---
title: "Bacterial bioluminescence inhibition (15 min) — drc analysis"
subtitle: "ISO 11348 · endpoint: relative light units"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
params:
  sample_id: "example"
  data_file: null
  test_type: "bioluminescence_inhibition"
  seed: 20260831
  compare_models: true
  fct: null
  save_fit: true
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(crworkflows)
library(dplyr)
library(ggplot2)

set.seed(params$seed)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5, dpi = 200)

# Quarto renders with the working directory set to this document, so the output
# root is resolved from the project marker rather than relatively. Without this
# every workflow would create its own outputs/ folder beside itself.
options(crworkflows.output_root = file.path(cr_project_root(), "outputs"))
```

## Purpose

This document runs a complete concentration-response analysis of a
Bacterial bioluminescence inhibition (15 min) test using the `drc` engine, and produces the figure, the estimate
table and the archived model fit that a report requires. It is intended to be
run once per sample. Change `params$sample_id` and `params$data_file` at the top
of the file; nothing below needs editing for a routine analysis.

The `drc` engine fits one mean function chosen from a small candidate set. That
choice is a decision made from the data, so it is shown in full below rather
than applied silently.

## Test and sample

```{r test-metadata}
tt <- cr_test_type(params$test_type)
knitr::kable(
  data.frame(
    Item = c("Sample", "Test type", "Guideline", "Endpoint",
             "Concentration units", "Error structure", "Default mean function"),
    Value = c(params$sample_id, tt$label, tt$guideline, tt$endpoint,
              tt$conc_units, tt$drc_type, tt$drc_fct)
  )
)
```

## Data

The default is the simulated example dataset shipped with the package. Set
`params$data_file` to a csv exported from the laboratory information system to
analyse a real sample. The csv must carry the columns listed above;
`inst/extdata/bioluminescence_inhibition.csv` is a template of the expected layout.

```{r load-data}
cr_data <- if (is.null(params$data_file)) {
  get(params$test_type, envir = asNamespace("crworkflows"))
} else {
  readr::read_csv(params$data_file, show_col_types = FALSE) |> as.data.frame()
}

knitr::kable(head(cr_data, 10), caption = "First ten rows of the analysis dataset")
```

## Data checks

The checks below are the ones that decide whether the analysis is valid, not a
summary of the data. Anything reported here must be resolved or recorded in the
report before the estimates are used.

```{r checks}
checks <- check_cr_data(cr_data, params$test_type)
checks
```

```{r design}
knitr::kable(
  summarise_design(cr_data, params$test_type),
  digits = 3,
  caption = "Replication and mean response at each concentration"
)
```

```{r raw-plot}
#| fig-cap: "Observed data. Open symbols are controls, plotted one decade below the lowest tested concentration because a zero cannot be placed on a logarithmic axis."
plot_cr_data(cr_data, params$test_type)
```

## Choice of mean function

Each candidate is fitted and compared. AIC ranks the candidates; the lack-of-fit
p-value tests the selected curve against a model with a separate mean at each
concentration, and a small value means the curve does not describe the data
regardless of how it ranks against the other candidates.

Selection is by AIC, but a fit with a lack-of-fit p-value below 0.05 should not
be reported without comment even when it ranks first.

```{r compare}
comparison <- if (params$compare_models) {
  compare_drc_models(cr_data, params$test_type)
} else {
  NULL
}
if (!is.null(comparison)) {
  knitr::kable(comparison, digits = 4, row.names = FALSE,
    caption = "Candidate mean functions, ordered by AIC")
}
```

```{r select}
selected <- if (!is.null(params$fct)) {
  params$fct
} else if (!is.null(comparison)) {
  comparison$fct[which(comparison$converged)][1]
} else {
  tt$drc_fct
}
cat("Mean function used:", selected, "\n")
```

## Model fitting

```{r fit}
fit <- fit_cr_drc(
  cr_data, params$test_type,
  fct = drc_fct(selected),
  validate = FALSE
)
summary(fit)
```

### Residual check

Systematic curvature in the residuals means the mean function is wrong, which
AIC alone will not reveal when every candidate is wrong in the same way.

```{r residuals}
#| fig-cap: "Residuals against fitted values. A trend, or a fan shape, indicates that the mean function or the error structure is misspecified."
resid_df <- data.frame(fitted = fitted(fit), residual = residuals(fit))
ggplot(resid_df, aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = "Fitted value", y = "Residual") +
  theme_bw()
```

## Results

ECx values are referenced to the fitted control response, which is the
convention in ecotoxicological reporting and matches what the `bayesnec`
workflow reports, so the two engines answer the same question.

```{r results}
results <- cr_results_table(
  fit, params$test_type,
  sample_id = params$sample_id
)
knitr::kable(results[, c("estimate_type", "level", "estimate", "lower", "upper")],
  digits = 4,
  caption = paste0("Estimates in ", tt$conc_units,
                   ". Intervals are 95% delta-method confidence intervals.")
)
```

```{r fit-plot}
#| fig-cap: "Fitted curve with its 95% confidence interval."
plot_cr_fit(fit, cr_data, params$test_type)
```

## Outputs and provenance

```{r outputs}
stem <- paste(params$sample_id, params$test_type, "drc", sep = "_")
paths <- write_cr_outputs(
  fit = fit,
  plot = plot_cr_fit(fit, cr_data, params$test_type),
  results = results,
  stem = stem,
  save_fit = params$save_fit
)
knitr::kable(data.frame(output = names(paths), path = unname(paths)))
```

```{r provenance}
knitr::kable(t(cr_session_record(seed = params$seed, engine = "drc")),
  col.names = "Value",
  caption = "Software versions and seed."
)
```
