---
title: "Larval development (48 h) — bayesnec analysis"
subtitle: "ASTM E724 · endpoint: proportion normal larvae"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
params:
  sample_id: "example"
  data_file: null
  test_type: "larval_development"
  seed: 20260831
  chains: 4
  iter: 4000
  warmup: 2000
  model: null
  backend: null
  save_fit: true
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(crworkflows)
library(dplyr)
library(ggplot2)

set.seed(params$seed)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5, dpi = 200)

# Quarto renders with the working directory set to this document, so the output
# root is resolved from the project marker rather than relatively. Without this
# every workflow would create its own outputs/ folder beside itself.
options(crworkflows.output_root = file.path(cr_project_root(), "outputs"))
```

## Purpose

This document runs a complete concentration-response analysis of a
Larval development (48 h) test using the `bayesnec` engine, and produces the figure, the
estimate table and the archived model fit that a report requires. It is
intended to be run once per sample. Change `params$sample_id` and
`params$data_file` at the top of the file; nothing below needs editing for a
routine analysis.

Every decision the analysis makes that is not routine is printed rather than
applied silently, so that it appears in the rendered record.

## Test and sample

```{r test-metadata}
tt <- cr_test_type(params$test_type)
knitr::kable(
  data.frame(
    Item = c("Sample", "Test type", "Guideline", "Endpoint",
             "Concentration units", "Response distribution", "Candidate models"),
    Value = c(params$sample_id, tt$label, tt$guideline, tt$endpoint,
              tt$conc_units, tt$bnec_family,
              if (is.null(params$model)) tt$bnec_model else params$model)
  )
)
```

## Data

The default is the simulated example dataset shipped with the package. Set
`params$data_file` to a csv exported from the laboratory information system to
analyse a real sample. The csv must carry the columns listed in the table
above; `inst/extdata/larval_development.csv` is a template of the expected layout.

```{r load-data}
cr_data <- if (is.null(params$data_file)) {
  get(params$test_type, envir = asNamespace("crworkflows"))
} else {
  readr::read_csv(params$data_file, show_col_types = FALSE) |> as.data.frame()
}

knitr::kable(head(cr_data, 10), caption = "First ten rows of the analysis dataset")
```

## Data checks

The checks below are the ones that decide whether the analysis is valid, not a
summary of the data. Anything reported here must be resolved or recorded in the
report before the estimates are used.

```{r checks}
checks <- check_cr_data(cr_data, params$test_type)
checks
```

```{r design}
knitr::kable(
  summarise_design(cr_data, params$test_type),
  digits = 3,
  caption = "Replication and mean response at each concentration"
)
```

```{r raw-plot}
#| fig-cap: "Observed data. Open symbols are controls, plotted one decade below the lowest tested concentration because a zero cannot be placed on a logarithmic axis."
plot_cr_data(cr_data, params$test_type)
```

## Model fitting

`bayesnec` fits every model in the candidate set and averages them by their
stacking weights, so no single model form has to be chosen in advance. The
family and its link come from the test-type registry: the link is always the
identity, which keeps `top`, `bot` and `nec` on the scale of the measured
response.

`params$backend` selects the Stan interface. Leaving it unset uses the `brms`
default, `rstan`. Setting it to `"cmdstanr"` uses CmdStan instead, which is the
faster of the two. Both compile Stan models and so both need a C++ toolchain.
The two are different implementations of the same sampler and are not
guaranteed to reproduce each other draw for draw from the same seed, so the
backend is recorded with the other provenance below.

```{r fit}
#| results: hide
fit_args <- list(
  cr_data, params$test_type,
  model = params$model,
  validate = FALSE,
  seed = params$seed,
  chains = params$chains,
  iter = params$iter,
  warmup = params$warmup,
  refresh = 0
)
if (!is.null(params$backend)) fit_args$backend <- params$backend
fit <- do.call(fit_cr_bayesnec, fit_args)
```

### Convergence

A fit that has not converged must not be reported. R-hat above 1.05, or a bulk
effective sample size below about 400, means the chains have not mixed and the
estimates below are not usable.

```{r convergence}
summary(fit)
```

### Model weights

Where more than one model was fitted, the weights show how much each
contributed to the averaged prediction. A weight concentrated entirely on one
model means the averaging has made no difference; weights spread across models
with different shapes mean the data do not determine the curve form, which is
worth recording in the report.

```{r weights}
if (inherits(fit, "bayesmanecfit")) {
  knitr::kable(cr_model_weights(fit), digits = 3, row.names = FALSE,
    caption = "Stacking weights, ordered by contribution")
} else {
  cat("A single model was fitted:", fit$model, "\n")
}
```

## Results

The threshold estimate is labelled with the quantity actually estimated.
`NEC` means every model in the fit carries the threshold parameter. `NSEC`
means none does, so the estimate comes from a smooth curve. `N(S)EC` means the
candidate set mixes the two and the model-averaged estimate is a weighted
mixture of the two kinds of draw; it is not a NEC and must not be reported as
one.

```{r results}
results <- cr_results_table(
  fit, params$test_type,
  sample_id = params$sample_id
)
knitr::kable(results[, c("estimate_type", "level", "estimate", "lower", "upper")],
  digits = 4,
  caption = paste0("Estimates in ", tt$conc_units, ". Intervals are 95% credible intervals.")
)
```

```{r fit-plot}
#| fig-cap: "Fitted model-averaged curve with its 95% credible interval."
plot_cr_fit(fit, cr_data, params$test_type)
```

## Outputs and provenance

```{r outputs}
stem <- paste(params$sample_id, params$test_type, "bayesnec", sep = "_")
paths <- write_cr_outputs(
  fit = fit,
  plot = plot_cr_fit(fit, cr_data, params$test_type),
  results = results,
  stem = stem,
  save_fit = params$save_fit
)
knitr::kable(data.frame(output = names(paths), path = unname(paths)))
```

```{r provenance}
knitr::kable(t(cr_session_record(seed = params$seed, engine = "bayesnec",
                                 backend = params$backend)),
  col.names = "Value",
  caption = "Software versions and seed. All of these are needed to reproduce the fit; the seed alone is not sufficient."
)
```
