---
title: "Earthworm acute survival (14 d) — drc analysis"
subtitle: "OECD 207 · endpoint: surviving adults"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
params:
  sample_id: "example"
  data_file: null
  test_type: "earthworm_survival"
  seed: 20260831
  average: true
  fct: null
  save_fit: true
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(crworkflows)
library(ggplot2)

set.seed(params$seed)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5, dpi = 200)

# Quarto renders with the working directory set to this document, so the output
# root is resolved from the project marker rather than relatively. Without this
# every workflow would create its own outputs/ folder beside itself.
options(crworkflows.output_root = file.path(cr_output_root(), "outputs"))
```

## Purpose

This document runs a complete concentration-response analysis of a
Earthworm acute survival (14 d) test using the `drc` engine, and produces the figure, the estimate
table and the archived model fit that a report requires. It is intended to be
run once per sample. Change `params$sample_id` and `params$data_file` at the top
of the file; nothing below needs editing for a routine analysis.

The `drc` engine fits a small set of candidate mean functions and, by default,
combines them by their Akaike weights. How the candidates were used, and how
much each contributed, is shown in full below rather than applied silently.

## Test and sample

```{r test-metadata}
tt <- cr_test_type(params$test_type)
knitr::kable(
  data.frame(
    Item = c("Sample", "Test type", "Guideline", "Endpoint",
             "Concentration units", "Error structure", "Default mean function"),
    Value = c(params$sample_id, tt$label, tt$guideline, tt$endpoint,
              tt$conc_units, tt$drc_type, tt$drc_fct)
  )
)
```

## Data

The default is the simulated example dataset shipped with the package. Set
`params$data_file` to a csv exported from the laboratory information system to
analyse a real sample. The csv must carry the columns listed above. A template
of the expected layout ships with the package; its path on this machine is
printed below.

```{r load-data}
cat("input format template:",
    system.file("extdata", "earthworm_survival.csv", package = "crworkflows"), "\n")

cr_data <- if (is.null(params$data_file)) {
  get(params$test_type, envir = asNamespace("crworkflows"))
} else {
  utils::read.csv(params$data_file, stringsAsFactors = FALSE)
}

knitr::kable(head(cr_data, 10), caption = "First ten rows of the analysis dataset")
```

## Data checks

The checks below are the ones that decide whether the analysis is valid, not a
summary of the data. Anything reported here must be resolved or recorded in the
report before the estimates are used.

```{r checks}
checks <- check_cr_data(cr_data, params$test_type)
checks
```

```{r design}
knitr::kable(
  summarise_design(cr_data, params$test_type),
  digits = 3,
  caption = "Replication and mean response at each concentration"
)
```

```{r raw-plot}
#| fig-cap: "Observed data. Open symbols are controls, plotted one decade below the lowest tested concentration because a zero cannot be placed on a logarithmic axis."
plot_cr_data(cr_data, params$test_type)
```

## Candidate mean functions

Each candidate is fitted and compared. AIC ranks them; the lack-of-fit p-value
tests a curve against a model with a separate mean at each concentration, and a
small value means the curve does not describe the data regardless of how it
ranks against the others. A candidate with a lack-of-fit p-value below 0.05
should not be reported without comment even when it ranks first.

```{r compare}
comparison <- compare_drc_models(cr_data, params$test_type)
knitr::kable(comparison, digits = 4, row.names = FALSE,
  caption = "Candidate mean functions, ordered by AIC")
```

## Model fitting

`params$average` controls how the candidates are used.

When it is `true`, which is the default, the candidates are combined by their
Akaike weights rather than one being chosen. The reported interval then includes
the uncertainty in the choice of curve form, instead of being conditional on one
function having been picked correctly. Where one candidate carries almost all
the weight the averaged result is indistinguishable from that candidate alone;
where the candidates disagree, the interval is wider, and that widening is the
information averaging adds. This matches what the `bayesnec` workflow does with
stacking weights, so the two engines report comparable quantities.

When it is `false`, the single best candidate by AIC is fitted and reported, and
the interval is conditional on that choice.

```{r fit}
selected <- if (is.null(params$fct)) {
  comparison$fct[which(comparison$converged)][1]
} else {
  params$fct
}

fit <- if (params$average) {
  fit_cr_drc_ma(cr_data, params$test_type, validate = FALSE)
} else {
  fit_cr_drc(cr_data, params$test_type, fct = drc_fct(selected), validate = FALSE)
}
```

```{r weights}
if (inherits(fit, "cr_drc_ma")) {
  knitr::kable(cr_model_weights(fit), digits = 4, row.names = FALSE,
    caption = "Akaike weights. These say how much each candidate contributed to the averaged result. They are not probabilities that a model is correct.")
} else {
  cat("Single mean function used:", selected, "\n\n")
  summary(fit)
}
```

### Residual check

Systematic curvature in the residuals means the mean function is wrong, which
AIC alone will not reveal when every candidate is wrong in the same way. For an
averaged fit the residuals of the highest-weighted candidate are shown, since
the averaged curve is a combination rather than a single fitted model.

```{r residuals}
#| fig-cap: "Residuals against fitted values. A trend, or a fan shape, indicates that the mean function or the error structure is misspecified."
resid_fit <- if (inherits(fit, "cr_drc_ma")) fit$fits[[fit$comparison$model[1]]] else fit
resid_df <- data.frame(fitted = fitted(resid_fit), residual = residuals(resid_fit))
ggplot(resid_df, aes(fitted, residual)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    x = "Fitted value", y = "Residual",
    subtitle = if (inherits(fit, "cr_drc_ma")) {
      paste("Highest-weighted candidate:", fit$comparison$model[1])
    } else {
      NULL
    }
  ) +
  theme_bw()
```

## Results

ECx values are referenced to the fitted control response, which is the
convention in ecotoxicological reporting and matches what the `bayesnec`
workflow reports, so the two engines answer the same question.

Where the candidates were averaged, the interval is a Buckland interval: each
candidate contributes its own uncertainty plus its distance from the averaged
value, so disagreement between curve forms widens it. Where a single candidate
was fitted, the interval is a delta-method confidence interval conditional on
that choice. The `interval` column of the results table records which was used.

```{r results}
results <- cr_results_table(
  fit, params$test_type,
  sample_id = params$sample_id
)
knitr::kable(results[, c("estimate_type", "level", "estimate", "lower", "upper")],
  digits = 4,
  caption = paste0("Estimates in ", tt$conc_units, ". ", results$interval[1])
)
```

```{r fit-plot}
#| fig-cap: "Fitted curve with its 95% confidence interval."
fit_plot <- plot_cr_fit(fit, cr_data, params$test_type)
fit_plot
```

## Outputs and provenance

```{r outputs}
stem <- paste(params$sample_id, params$test_type, "drc", sep = "_")
paths <- write_cr_outputs(
  fit = fit,
  plot = fit_plot,
  results = results,
  stem = stem,
  save_fit = params$save_fit
)
knitr::kable(data.frame(output = names(paths), path = unname(paths)))
```

```{r provenance}
knitr::kable(t(cr_session_record(seed = params$seed, engine = "drc")),
  col.names = "Value",
  caption = "Software versions and seed."
)
```
